---
title: "Certifying outputs and detecting drift"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Certifying outputs and detecting drift}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(reproducr)
cert_file <- tempfile()   # shared across all chunks
```

This vignette covers Tier 2 of the `reproducr` workflow in depth:
`certify()`, `check_drift()`, and `list_certs()`. These three functions
together form the *baseline and drift detection* system.

## The problem they solve

`renv` freezes the package versions you declare. But packages change hands,
maintainers push silent fixes, platform-level libraries (BLAS, LAPACK) get
updated by system administrators, and R itself changes RNG defaults between
minor versions. Any of these can alter your numerical results without
producing an error.

`certify()` and `check_drift()` detect this. The idea is simple:

1. After a successful analysis run, hash the key outputs and store the hashes.
2. Later — after any change to the environment — re-run the analysis and
   compare the new hashes against the stored ones.
3. Any mismatch is reported explicitly, by output name.

---

## `certify()` — creating a baseline

### What gets hashed

Pass a fully named list of any R objects you want to protect. Common choices:

```{r certify-examples}
model <- lm(mpg ~ wt + cyl, data = mtcars)

certify(
  outputs = list(
    # Model parameters
    coefs       = coef(model),
    r_squared   = summary(model)$r.squared,
    sigma       = sigma(model),

    # Key data properties
    n_obs       = nrow(mtcars),
    n_complete  = sum(complete.cases(mtcars)),

    # A summary table
    group_means = aggregate(mpg ~ cyl, data = mtcars, FUN = mean)
  ),
  tag    = "baseline-v1",
  script = "analysis.R",
  file   = cert_file
)
```

Any R object that can be serialised works: numeric vectors, data frames,
matrices, lists, character vectors, and scalars. The hash is computed with
SHA-256 via the `digest` package if available, or a base-R fallback otherwise.

### Choosing what to certify

Certify outputs that are:

- **Conclusions** — the numbers that appear in your paper or report
- **Stable** — not random session artefacts like timestamps or row ordering
- **Interpretable** — so a drift report tells you something meaningful

Avoid certifying objects that are expected to differ across runs by design,
such as `proc.time()` outputs or `Sys.time()` values.

### Tags and the certification store

Every certification requires a `tag` — a human-readable label that identifies
the analysis milestone:

```{r tags}
# Certify at multiple milestones
certify(
  outputs = list(coefs = coef(model)),
  tag     = "pre-peer-review",
  file    = cert_file
)

certify(
  outputs = list(coefs = coef(model)),
  tag     = "post-revision",
  file    = cert_file
)
```

All certifications for a project accumulate in a single `.reproducr.rds` file.
Tags must be unique within a file — passing a duplicate tag overwrites the
existing record with a warning:

```{r duplicate-tag, warning = TRUE}
certify(
  outputs = list(coefs = coef(model)),
  tag     = "baseline-v1",  # already exists
  file    = cert_file
)
```

### What gets stored alongside the hashes

Each certification record stores:

| Field | Description |
|---|---|
| `tag` | The label you supplied |
| `timestamp` | ISO 8601 timestamp |
| `script` | Path to the script (if supplied) |
| `r_version` | R version at certification time |
| `r_platform` | Platform string |
| `os` | Operating system |
| `hashes` | Named list of SHA-256 hashes |
| `output_names` | Names of certified outputs |
| `output_classes` | R class of each output |
| `output_dims` | Dimensions/length of each output |

This means that even if `check_drift()` reports a mismatch, the record tells
you what *type* of object changed and approximately what size it was — which
helps diagnose whether it was a data issue or a model issue.

---

## `list_certs()` — inspecting the store

Before running `check_drift()`, you can inspect what certifications are stored:

```{r list-certs}
list_certs(file = cert_file)
```

The result is a plain data frame — one row per certification, in insertion
order. This is useful for:

- Confirming the right tag exists before passing it to `check_drift()`
- Auditing the history of an analysis over time
- Checking that CI has been certifying on each run

---

## `check_drift()` — comparing against a baseline

### Basic usage

```{r drift-basic}
# Re-run the same analysis
model2 <- lm(mpg ~ wt + cyl, data = mtcars)

result <- check_drift(
  outputs = list(
    coefs      = coef(model2),
    r_squared  = summary(model2)$r.squared,
    sigma      = sigma(model2),
    n_obs      = nrow(mtcars),
    n_complete = sum(complete.cases(mtcars)),
    group_means = aggregate(mpg ~ cyl, data = mtcars, FUN = mean)
  ),
  against = "baseline-v1",
  file    = cert_file
)
```

### The four statuses

Every output in the comparison gets one of four statuses:

```{r statuses}
# Demonstrate all four by constructing an artificial scenario
certify(
  outputs = list(
    stays_same  = 42L,
    will_change = coef(lm(mpg ~ wt, data = mtcars)),
    will_vanish = "this output disappears next run"
  ),
  tag  = "four-statuses",
  file = cert_file
)

demo_result <- check_drift(
  outputs = list(
    stays_same  = 42L,
    will_change = coef(lm(mpg ~ hp, data = mtcars)),  # different model
    brand_new   = "this output is new"
    # will_vanish is not supplied
  ),
  against = "four-statuses",
  file    = cert_file
)

print(demo_result)
```

| Status | Meaning |
|---|---|
| `ok` | Hash matches the baseline exactly |
| `drifted` | Hash differs — output has changed |
| `missing` | Present in baseline, not supplied to `check_drift()` |
| `new` | Supplied to `check_drift()`, not in baseline |

`"missing"` and `"new"` are not errors by themselves — you may have
intentionally added or removed outputs between milestones. But they are worth
reviewing, especially `"missing"`, which could indicate an output was dropped
by accident.

### Using `"latest"` to always compare against the most recent certification

```{r latest}
certify(outputs = list(x = 1L), tag = "run-1", file = cert_file)
certify(outputs = list(x = 1L), tag = "run-2", file = cert_file)
certify(outputs = list(x = 1L), tag = "run-3", file = cert_file)

# Compares against "run-3" automatically
check_drift(outputs = list(x = 1L), against = "latest", file = cert_file)
```

This is the pattern to use in CI: always certify after a successful run, then
on the next run compare against `"latest"`. Any drift between consecutive runs
is caught immediately.

### Working with the drift report

The return value is a `data.frame` with columns `output`, `status`,
`max_delta`, and `note`. Use it programmatically to make CI fail on drift:

```{r drift-programmatic, eval = FALSE}
result <- check_drift(outputs = current_outputs, against = "latest")

n_drifted <- sum(result$status == "drifted")
if (n_drifted > 0L) {
  drifted_names <- result$output[result$status == "drifted"]
  stop(sprintf(
    "%d output(s) have drifted since last certification: %s",
    n_drifted,
    paste(drifted_names, collapse = ", ")
  ))
}
```

---

## Recommended workflow across a project lifecycle

### Initial analysis

```r
# Run analysis
model <- lm(...)
results <- compute_results(...)

# Certify at submission
certify(
  outputs = list(
    primary_coef = coef(model)[2],
    primary_pval = summary(model)$coefficients[2, 4],
    n            = nrow(data),
    effect_size  = compute_d(model)
  ),
  tag    = "submitted-2026-01-15",
  script = "main_analysis.R"
)
```

### After reviewer comments

```r
# Re-run with minor changes
# ...

check_drift(
  outputs = list(
    primary_coef = coef(model)[2],
    primary_pval = summary(model)$coefficients[2, 4],
    n            = nrow(data),
    effect_size  = compute_d(model)
  ),
  against = "submitted-2026-01-15"
)
# If all "ok" — results are numerically identical to submission
# If "drifted" — investigate before resubmitting
```

### In a CI pipeline (GitHub Actions)

```yaml
- name: Certify and check drift
  run: |
    Rscript -e "
      source('main_analysis.R')
      library(reproducr)

      # On first run, certify
      if (!file.exists('.reproducr.rds')) {
        certify(outputs = OUTPUTS, tag = 'ci-initial')
      }

      # On every subsequent run, check
      result <- check_drift(outputs = OUTPUTS, against = 'latest')
      certify(outputs = OUTPUTS, tag = paste0('ci-', Sys.Date()))

      if (any(result\$status == 'drifted')) stop('Drift detected!')
    "
```

---

## Version control considerations

Commit `.reproducr.rds` to your Git repository. This gives you:

- A permanent, auditable history of what every run produced
- The ability to `check_drift()` against any past milestone, not just the
  most recent one
- A single file reviewers can inspect to verify reproducibility claims

Add `.reproducr.rds` to your `.gitattributes` as a binary file to prevent
noisy diffs:

```
.reproducr.rds binary
```

And add a note to your `.Rbuildignore` so it does not end up in a package
tarball if your analysis is inside a package directory:

```
^\.reproducr\.rds$
```

```{r cleanup, include = FALSE}
unlink(paste0(cert_file, ".rds"))
```
