---
title: "Contributing to the breaking-changes database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contributing to the breaking-changes database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(reproducr)
```

The breaking-changes database is the heart of `risk_score()`'s
`"changelog"` check. It is a curated list of cases where a package update
silently changed a function's *behaviour* — not its interface — in a way that
can alter analytical results without producing an error.

This vignette explains the database schema, how to find good candidates,
how to write an entry, and how to test it before submitting a pull request.

---

## What belongs in the database

A good database entry documents a case where all of the following are true:

1. **A specific function in a specific package** changed its output between two
   versions.
2. The change is **silent** — it does not throw an error or warning on the old
   calling pattern.
3. The change can **affect analytical conclusions** — not just cosmetic
   differences like whitespace in printed output.
4. The change is **documented** — there is a `NEWS.md`, GitHub release, or
   official changelog entry you can link to.

### Examples of things that belong

- `dplyr::summarise()` changed its default `.groups` argument in v1.1.0,
  causing grouped data frames to be returned ungrouped where they previously
  were not.
- `stats::sample()` RNG algorithm changed in R 3.6.0, producing different
  random draws for the same seed.
- `readr::read_csv()` switched its parsing backend in v2.0.0, changing column
  type inference.

### Examples of things that do not belong

- A function that **errors** on the old calling pattern after an upgrade
  (that's a breaking change that produces a visible failure, not a silent one).
- A cosmetic change to printed output (e.g. number of decimal places shown by
  `print.lm()`).
- A change that only affects internal implementation with no observable
  difference in returned values.
- Performance changes with no effect on results.

---

## The database schema

The database lives in `R/breaking_changes_db.R` as a named list called
`.BREAKING_CHANGES_DB`. Keys are `"pkg::fn"` strings. Each key maps to a
list of entries — one per distinct breaking change for that function:

```r
.BREAKING_CHANGES_DB <- list(

  "pkg::function_name" = list(
    list(
      from_version = "x.y.z",   # last version where behaviour was "safe"
      to_version   = "x.y.z",   # last version where breaking change applies
      risk         = "high",     # "high", "medium", or "low"
      description  = "...",      # plain-English explanation
      reference    = "https://..." # URL to the changelog entry
    )
  )

)
```

### The version window

The `from_version` and `to_version` fields define a **half-open interval**:

```
(from_version, to_version]
     ↑               ↑
 exclusive        inclusive
```

A user's installed version is flagged if and only if:

```
installed > from_version  AND  installed <= to_version
```

**`from_version`** is the last version where the old behaviour still applied.
Set this to one patch version below the first risky version. For example, if
the breaking change was introduced in `1.1.0`, set `from_version = "1.0.99"`.

**`to_version`** is the last version where the breaking change is known to
apply. If the behaviour has since been fixed or stabilised, set this to the
last affected version. If the change is permanent and no fix has been released,
set a high ceiling like `"9.9.9"` to cover all future versions.

### Risk levels

| Level | When to use |
|---|---|
| `"high"` | Output *values* change silently — model coefficients, table cells, random draws, sort order. Any result that goes into a paper could be different. |
| `"medium"` | An argument was renamed or deprecated; the function may warn, error, or produce different output depending on the call pattern. Needs manual review. |
| `"low"` | A minor behavioural note. Output is unlikely to differ in practice, but worth knowing. Covers locale sensitivity, cosmetic differences in edge cases, etc. |

---

## Finding candidates

### From NEWS.md files

Most CRAN packages maintain a `NEWS.md`. Look for entries mentioning:

- "breaking change"
- "changed default"
- "deprecated"
- "now returns"
- "behaviour changed"
- "no longer"

The tidyverse packages maintain especially detailed changelogs:

- `dplyr`: <https://dplyr.tidyverse.org/news/index.html>
- `tidyr`: <https://tidyr.tidyverse.org/news/index.html>
- `ggplot2`: <https://ggplot2.tidyverse.org/news/index.html>
- `readr`: <https://readr.tidyverse.org/news/index.html>
- `purrr`: <https://purrr.tidyverse.org/news/index.html>
- `stringr`: <https://stringr.tidyverse.org/news/index.html>
- `lubridate`: <https://lubridate.tidyverse.org/news/index.html>

For base R, the R release notes are authoritative:

- <https://cran.r-project.org/doc/manuals/r-release/NEWS.html>

### From your own experience

If you have encountered a case where upgrading a package changed your results,
that is exactly the kind of entry the database needs. Document it while the
details are fresh.

---

## Writing an entry

Here is a complete worked example. Suppose you have discovered that
`broom::augment()` changed the name of the residuals column from `.resid`
to `.std.resid` for some model types between versions 0.7.x and 0.8.0.

**Step 1: Verify it is in the changelog.**

Find the relevant `NEWS.md` entry. Copy the URL.

**Step 2: Determine the version window.**

The change was introduced in 0.8.0. The last safe version is 0.7.x. Set:
- `from_version = "0.7.99"` (one patch below the first risky version)
- `to_version = "1.0.9"` (the last version where this applies, or a high
  ceiling if unfixed)

**Step 3: Determine the risk level.**

Code that selects the `.resid` column by name from `augment()` output will
silently return `NA` or error after the upgrade, depending on how selection
is done. Categorise as `"medium"` — the calling pattern may break, but it
depends on the code.

**Step 4: Write the description.**

The description should:
- State which version the change was introduced in
- Describe the old behaviour
- Describe the new behaviour
- Explain the practical consequence for reproducibility

**Step 5: Add the entry.**

Open `R/breaking_changes_db.R` and add your entry in the appropriate section:

```r
"broom::augment" = list(
  list(
    from_version = "0.7.99",
    to_version   = "1.0.9",
    risk         = "medium",
    description  = paste0(
      "In broom 0.8.0, augment() renamed residual columns for several model ",
      "types — '.resid' became '.std.resid' for standardised residuals in ",
      "lm and glm objects. Code that selects '.resid' by name from augment() ",
      "output will silently return NA or error after upgrading."
    ),
    reference = "https://broom.tidymodels.org/news/index.html"
  )
),
```

---

## Testing your entry

Before submitting, verify that the entry behaves correctly by simulating a
version match:

```{r test-entry}
# Temporarily add a test entry to verify the detection logic
# (this is what a test in tests/testthat/ would do)

# 1. Write a script that calls the at-risk function
test_script <- tempfile(fileext = ".R")
writeLines("x <- dplyr::summarise(mtcars, n = dplyr::n())", test_script)

# 2. Audit it — package version will be from installed library
report <- audit_script(test_script, renv = FALSE, verbose = FALSE)

# 3. Confirm the call was detected
report$calls[, c("pkg", "fn", "pkg_version")]

# 4. Run the changelog check
risks <- risk_score(report, methods = "changelog")

# 5. If the installed version is in the risk window, it will be flagged
if (nrow(risks) > 0) {
  as.data.frame(risks)[, c("call", "pkg_version", "risk", "check")]
} else {
  cat("Installed version is outside the risk window — entry is correct,\n")
  cat("but not triggered by this version.\n")
}
```

To test with a specific version (simulating a user on an older installation),
you can temporarily override the `pkg_version` in the calls data frame:

```{r test-with-version}
# Force the version to sit inside the dplyr::summarise risk window
report_forced <- report
report_forced$calls$pkg_version <- as.character(report_forced$calls$pkg_version)
report_forced$calls$pkg_version[report_forced$calls$pkg == "dplyr"] <- "1.1.0"

risks_forced <- risk_score(report_forced, methods = "changelog")
as.data.frame(risks_forced)[, c("call", "pkg_version", "risk")]
```

---

## Submitting a pull request

1. Fork the repository on GitHub.
2. Add your entry to `R/breaking_changes_db.R` in the appropriate package
   section (alphabetical order by package name, then by function name).
3. Add a test in `tests/testthat/test-risk_score.R` that confirms:
   - The entry is flagged when the version is inside the risk window.
   - The entry is **not** flagged when the version is outside the window.
4. Update `NEWS.md` with a bullet under `# reproducr (development version)`.
5. Open a pull request with a short description of the breaking change and
   a link to the official changelog.

### Minimum test for a new entry

```r
test_that("risk_score() flags mypackage::myfun in the risk window", {
  f <- write_script("x <- mypackage::myfun(42)")
  on.exit(unlink(f))
  r <- audit_script(f, renv = FALSE, verbose = FALSE)

  # Simulate a version inside the risk window
  r$calls$pkg_version[r$calls$pkg == "mypackage"] <- "2.1.0"

  rs <- risk_score(r, methods = "changelog")
  expect_true(nrow(rs) > 0L)
  expect_true(any(rs$call == "mypackage::myfun"))
})

test_that("risk_score() does NOT flag mypackage::myfun outside the window", {
  f <- write_script("x <- mypackage::myfun(42)")
  on.exit(unlink(f))
  r <- audit_script(f, renv = FALSE, verbose = FALSE)

  # Simulate a version before the risk window
  r$calls$pkg_version[r$calls$pkg == "mypackage"] <- "1.9.0"

  rs <- risk_score(r, methods = "changelog")
  myfun_risks <- rs[rs$call == "mypackage::myfun" & rs$check == "changelog", ]
  expect_equal(nrow(myfun_risks), 0L)
})
```

```{r cleanup, include = FALSE}
unlink(test_script)
```
