---
title: "Getting started with reproducr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with reproducr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = TRUE
)
library(reproducr)
```

## What is reproducr?

`reproducr` audits R scripts for computational reproducibility risk — filling
the gap that `renv` leaves. While `renv` is excellent at freezing package
versions, it cannot detect whether a function's *behaviour* changed silently
between versions, whether stochastic code is missing a `set.seed()`, or
whether analytical results have numerically drifted after a package upgrade.

`reproducr` answers those questions via a three-tier workflow:

1. **Scan & score** — parse your scripts and assess risk
2. **Baseline & drift** — certify outputs and detect changes over time
3. **Report & export** — generate human-readable audit reports

## Tier 1: Scan and score

### Auditing a script

The entry point is `audit_script()`. It reads your R source files, extracts
every qualified `pkg::fn` call, and resolves which version of each package is
in use (from `renv.lock` if available, otherwise the installed library).

```{r audit}
# Create a small example script
script <- tempfile(fileext = ".R")
writeLines(c(
  "# Example analysis",
  "set.seed(42)",
  "x   <- dplyr::filter(mtcars, cyl == 4)",
  "y   <- dplyr::summarise(x, mean_mpg = mean(mpg), n = dplyr::n())",
  "fit <- lm(mpg ~ wt, data = x)",
  "z   <- stats::rnorm(nrow(y))",
  "out <- base::sort(unique(x$gear))"
), script)

report <- audit_script(script, renv = FALSE, verbose = FALSE)
print(report)
```

The `audit_report` object holds the detected calls as a data frame:

```{r calls}
report$calls
```

### Scoring for risk

Pass the report to `risk_score()` to run three independent checks:

```{r risk}
risks <- risk_score(report)
print(risks)
```

The `"changelog"` check compares each call against a curated database of
known breaking changes. The `"seed_check"` verifies stochastic functions have
`set.seed()` nearby. The `"locale_check"` flags locale-sensitive operations.

You can run individual checks or filter by minimum risk level:

```{r risk-filtered}
# High-severity only
high_risks <- risk_score(report, min_risk = "high")

# Just the seed check
seed_issues <- risk_score(report, methods = "seed_check")
```

The result is a plain data frame, so you can filter, sort, and export it with
standard R tools:

```{r risk-df}
# All risks as a regular data frame
as.data.frame(risks)
```

## Tier 2: Baseline and drift detection

### Certifying outputs

After running an analysis, certify the key outputs using `certify()`. This
hashes each object and stores the hash alongside full environment metadata.

```{r certify}
cert_file <- tempfile()  # Use ".reproducr" in a real project

model <- lm(mpg ~ wt, data = mtcars)

certify(
  outputs = list(
    coefs     = coef(model),
    r_squared = summary(model)$r.squared,
    n_obs     = nrow(mtcars)
  ),
  tag    = "baseline-v1",
  script = script,
  file   = cert_file
)
```

Inspect stored certifications with `list_certs()`:

```{r list-certs}
list_certs(file = cert_file)
```

### Checking for drift

After a package upgrade or environment change, re-run `check_drift()` to
verify that your results have not changed:

```{r drift-ok}
# Same outputs — should report "ok"
result <- check_drift(
  outputs = list(
    coefs     = coef(model),
    r_squared = summary(model)$r.squared,
    n_obs     = nrow(mtcars)
  ),
  against = "baseline-v1",
  file    = cert_file
)
```

If an output changes, `check_drift()` reports it clearly:

```{r drift-detect}
# Different model — will show drift
model2 <- lm(mpg ~ hp, data = mtcars)

check_drift(
  outputs = list(coefs = coef(model2)),
  against = "baseline-v1",
  file    = cert_file
)
```

The drift report is also a data frame you can inspect:

```{r drift-df}
print(result)
```

## Tier 3: Report and export

### Generating reports

`repro_report()` renders audit findings in three styles:

**Minimal** — a compact summary:

```{r report-minimal, results='hide'}
repro_report(report, risks, format = "text", style = "minimal")
```

**Academic** — a methods paragraph ready to paste into a paper:

```{r report-academic}
cat(repro_report(report, risks, format = "text", style = "academic"))
```

**Pharma** — a structured QC document with risk register and sign-off fields:

```{r report-pharma, results='hide'}
html_file <- tempfile(fileext = ".html")
repro_report(report, risks,
             format      = "html",
             style       = "pharma",
             output_file = html_file)
```

### Generating badges

`repro_badge()` creates a shields.io badge reflecting the current
reproducibility status:

```{r badge}
badge <- repro_badge(report, risks, output = "markdown")
cat(badge)
```

To insert the badge into your `README.md` automatically (e.g. from CI):

```r
repro_badge(report, risks, output = "README")
```

## The full pipeline

```r
library(reproducr)

# Tier 1
report <- audit_script("analysis.R")
risks  <- risk_score(report)

# Tier 2
certify(
  outputs = list(coefs = coef(my_model)),
  tag     = "submission-v1"
)

# After an upgrade:
check_drift(
  outputs = list(coefs = coef(my_model)),
  against = "submission-v1"
)

# Tier 3
repro_report(report, risks, format = "html", style = "pharma")
repro_badge(report, risks, output = "README")
```

## Working with renv

`reproducr` is designed to complement `renv`, not replace it. When an
`renv.lock` file is present, `audit_script()` reads package versions from
the lockfile automatically:

```r
# renv.lock present in working directory — versions read automatically
report <- audit_script("analysis.R")  # renv = TRUE by default
```

Commit both `renv.lock` and `.reproducr.rds` to version control for a
complete, auditable reproducibility record.

```{r cleanup, include = FALSE}
unlink(script)
unlink(paste0(cert_file, ".rds"))
```
