---
title: "Getting started with reproducr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with reproducr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = TRUE
)
library(reproducr)
```

## What is reproducr?

You finish an analysis. The code runs. The numbers look right. But are they stable?

Package updates change function behaviour silently. Stochastic code without a fixed seed
produces different results on every run. Results certified last month may drift this month
— with no error and no warning.

`reproducr` makes these risks visible and trackable via a three-tier workflow:

1. **Scan & score** — parse your scripts and assess risk
2. **Baseline & drift** — certify outputs and detect changes over time
3. **Report & export** — generate human-readable audit reports

It works with your existing setup. If you use `renv`, `reproducr` reads your lockfile
automatically. No configuration required.

---

## Why this matters — real failure modes

These are not hypothetical. Each scenario describes a class of problem that occurs
routinely in research and regulated workflows, produces no error, and is invisible without
explicit tooling.

### Scenario 1 — The collaborator upgrade problem

You write an analysis in January using dplyr 1.0.4 and share it with a colleague who has
dplyr 1.1.2.

```r
results <- mtcars |>
  dplyr::group_by(cyl) |>
  dplyr::summarise(mean_mpg = mean(mpg))

# You then chain a further operation:
results |> dplyr::mutate(rank = dplyr::row_number())
```

In dplyr 1.0.x, `summarise()` retained grouping by default. In dplyr 1.1.x it drops the
last grouping level. Your colleague's `mutate()` now operates on ungrouped data — the
`rank` column is computed differently. No error. No warning. Different numbers.

`reproducr` flags this immediately:

```
[HIGH] dplyr::summarise
       In dplyr 1.1.0, summarise() changed its default grouping behaviour...
```

### Scenario 2 — The server deployment problem

You develop a model locally on R 3.5.3 and deploy to a production server running R 3.6.2.

```r
set.seed(42)
train_idx <- base::sample(1:nrow(data), 0.8 * nrow(data))
```

R 3.6.0 changed the default RNG algorithm for `sample()`. The same seed now produces a
different train/test split. Your model is trained on different data than you validated
locally. Accuracy metrics differ silently across environments.

`reproducr` flags this:

```
[HIGH] stats::sample
       In R 3.6.0, the default RNG algorithm changed...
```

### Scenario 3 — The renv false sense of security

You use `renv` to lock your environment and restore it six months later on a new machine.
Everything installs correctly but results differ.

`renv` locked `readr 2.0.1`. Your original analysis was written with `readr 1.4.0`. The
lockfile captured the version you were *already on* when you ran `renv::init()` — past
the breaking change. You never compared against pre-2.0 output.

```r
data <- readr::read_csv("clinical_data.csv")
# Column "patient_id" now parses as character instead of double.
# Downstream merge silently drops rows.
```

`renv` cannot detect this because it only sees versions, not behaviour. `reproducr` sees
the function call and flags it:

```
[HIGH] readr::read_csv
       In readr 2.0.0, read_csv() switched to the vroom backend.
       Column type guessing changed...
```

---

## Tier 1: Scan and score

### Auditing a script

The entry point is `audit_script()`. It reads your R source files, extracts every
qualified `pkg::fn` call, and resolves which version of each package is in use.

```{r audit}
# Create a small example script
script <- tempfile(fileext = ".R")
writeLines(c(
  "# Example analysis",
  "set.seed(237)",
  "x   <- dplyr::filter(mtcars, cyl == 4)",
  "y   <- dplyr::summarise(x, mean_mpg = mean(mpg), n = dplyr::n())",
  "fit <- lm(mpg ~ wt, data = x)",
  "z   <- stats::rnorm(nrow(y))",
  "out <- base::sort(unique(x$gear))"
), script)

report <- audit_script(script, renv = FALSE, verbose = FALSE)
print(report)
```

```{r calls}
report$calls
```

### Scoring for risk

Pass the report to `risk_score()` to run three independent checks:

```{r risk}
risks <- risk_score(report)
print(risks)
```

- **`"changelog"`** — checks calls against a curated database of known silent breaking changes
- **`"seed_check"`** — flags stochastic functions without a nearby `set.seed()`
- **`"locale_check"`** — flags functions whose output varies by system locale

```{r risk-filtered}
# High-severity only
high_risks <- risk_score(report, min_risk = "high")

# Just the seed check
seed_issues <- risk_score(report, methods = "seed_check")
```

```{r risk-df}
# As a plain data frame for downstream use
as.data.frame(risks)
```

## Tier 2: Baseline and drift detection

### Certifying outputs

After running an analysis, certify the key outputs using `certify()`.

```{r certify}
cert_file <- tempfile()

model <- lm(mpg ~ wt, data = mtcars)

certify(
  outputs = list(
    coefs     = coef(model),
    r_squared = summary(model)$r.squared,
    n_obs     = nrow(mtcars)
  ),
  tag    = "baseline-v1",
  script = script,
  file   = cert_file
)
```

```{r list-certs}
list_certs(file = cert_file)
```

### Checking for drift

After any environment change, re-run `check_drift()`:

```{r drift-ok}
result <- check_drift(
  outputs = list(
    coefs     = coef(model),
    r_squared = summary(model)$r.squared,
    n_obs     = nrow(mtcars)
  ),
  against = "baseline-v1",
  file    = cert_file
)
```

```{r drift-detect}
# Different model — shows drift
model2 <- lm(mpg ~ hp, data = mtcars)

check_drift(
  outputs = list(coefs = coef(model2)),
  against = "baseline-v1",
  file    = cert_file
)
```

## Tier 3: Report and export

```{r report-minimal, results='hide'}
repro_report(report, risks, format = "text", style = "minimal")
```

```{r report-academic}
cat(repro_report(report, risks, format = "text", style = "academic"))
```

```{r badge}
badge <- repro_badge(report, risks, output = "markdown")
cat(badge)
```

## The full pipeline

```r
library(reproducr)

# Tier 1
report <- audit_script("analysis.R")
risks  <- risk_score(report)

# Tier 2
certify(
  outputs = list(coefs = coef(my_model)),
  tag     = "submission-v1"
)

check_drift(
  outputs = list(coefs = coef(my_model)),
  against = "submission-v1"
)

# Tier 3
repro_report(report, risks, format = "html", style = "pharma")
repro_badge(report, risks, output = "README")
```

```{r cleanup, include = FALSE}
unlink(script)
unlink(paste0(cert_file, ".rds"))
```
